---
title: "Class 15: Pertussis Mini Project"
author: "Olivia Baldwin"
format: pdf
---

Pertussis (aka Whooping Cough) is a highly contagious lung infection cause by the bacteria *B. Pertussis*. 

The CDC tracks Pertussis case numbers and they can be accessed [here](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html). 

We need to "scrape" this data so we can do stuff with it in R. Let's use the **datapasta** package. 

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L, 2024L),
                                cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044, 23544)
       )
```

Let's plot the cases per year to see the trend over time in the US. 

```{r}
library(ggplot2)
```

```{r}
baseplot <- ggplot(cdc, aes(x=year, y=cases)) + 
  geom_point() +
  geom_line() 

baseplot
```

Let's add the date of the wP vaccine roll out completion (1946) and the new aP vaccine (1996). 

```{r}
baseplot +
  theme_bw() +
  geom_vline(xintercept = 1946, col="blue") +
  geom_vline(xintercept = 1996, col="red") +
  geom_vline(xintercept = 2020, col="grey") +
  geom_vline(xintercept = 2004, col= "forestgreen")
```

## CMI-PB (Computational Models of Immunity - Pertussis Boost)

This project collects and makes freely available data about the immune response to Pertussis vaccination. 

You can access the data via an API which returns a JSON format (key:value). 

We can use the **jsonlite** package and it's `read_json()` function. 

```{r}
library(jsonlite) 
```

```{r}
subject <- read_json("http://cmi-pb.org/api/v5/subject", simplifyVector = TRUE) 
#simplifyVector will read it in as a data frame rather than key:value 

head(subject)
```

>Q: How many subjects are there? **172**

```{r}
dim(subject) 
```

>Q: How many male? Female? **60 male, 112 female** 

```{r}
table(subject$biological_sex) 
```

>Q: How many wP and aP? **85 wP and 87 aP** 

```{r}
table(subject$infancy_vac)
```

>Q: Break down of biological sex and race? 

```{r}
table(subject$race, subject$biological_sex)  
```

>Q: Does this breakdown reflect the US population? **no**  

```{r}
table(subject$dataset) 
```

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5/specimen", simplifyVector = TRUE) 

ab_titer <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer", simplifyVector = TRUE)

head(specimen) 
head(ab_titer)
```

```{r}
library(dplyr) 
```

We want to merge these tables so we can have all the info we need about a given antibody measurement. 

The `inner_join()` function will keep only things in common between the two data frames when it combines them. `full_join()` would keep everything. 

```{r}
meta <- inner_join(subject, specimen) 
```

```{r}
head(meta)
```

Now we need to join the meta and the ab_titer to get everything in one place. 

```{r}
abdata <- inner_join(ab_titer, meta) 

head(abdata)
```

```{r}
nrow(abdata) 
```

```{r}
table(abdata$isotype) 
```

```{r}
table(abdata$antigen)
```

Let's begin with IgG

```{r}
igg <- filter(abdata, isotype=="IgG") 
head(igg) 
```

Make a boxplot of IgG antigen levels - this will be a plot of MFI normalized vs antigen 

```{r}
igg_plot <- ggplot(igg, aes(MFI_normalised, antigen)) + 
  geom_boxplot() 

igg_plot
```

```{r}
igg_plot + aes(col=infancy_vac)
```


Ideally I would like to see how these Ab levels change over time relative to the booster shot. 

```{r}
table(abdata$visit) 
```

```{r}
igg_plot + facet_wrap(~infancy_vac) 
```

```{r}
igg_plot + 
  aes(col=infancy_vac) + 
  facet_wrap(~visit) 
```

```{r}
head(igg)
```

```{r}
igg_21_PT <- filter(igg, antigen == "PT", dataset == "2021_dataset")
head(igg_21_PT)
```

```{r}
ggplot(igg_21_PT) +
  aes(actual_day_relative_to_boost, MFI_normalised, col=infancy_vac) +
  geom_point() +
  geom_line(aes(group=subject_id)) +
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_vline(xintercept = 14, linetype="dashed") +
  geom_smooth(se=FALSE) +
  labs(title="2021 IgG PT Response Over Time")
```

```{r}
igg_23_PT <- filter(igg, dataset=="2023_dataset", antigen=="PT") 
```

There is no data for after day 0 in this dataset...

```{r}
ggplot(igg_23_PT) +
  aes(actual_day_relative_to_boost, MFI_normalised, col=infancy_vac) +
  geom_point() +
  geom_line(aes(group=subject_id)) +
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_vline(xintercept = 14, linetype="dashed") +
  geom_smooth(se=FALSE) +
  labs(title="2023 IgG PT Response Over Time")
```


