---
title: "Class 9: Halloween Mini Project"
author: "Olivia Baldwin"
format: pdf
---

```{r}
candy <- read.csv("candy-data.csv", row.names = 1) 
head(candy)
```

```{r}
dim(candy)
table(candy$fruity)
```

>Q1: There are 85 different types of candy (rows)

>Q2: There are 38 "fruity" candies.


```{r}
candy["Kit Kat", ]$winpercent
candy["Tootsie Roll Snack Bars", ]$winpercent
```

>Q3: I like kit kats and their win percentage is 76.77% 

>Q4: Kit kat has a win percentage of 76.77%

>Q5: Tootsie roll snack bars have a win percentage of 49.65%


```{r}
library("skimr")
skim(candy)
```

>Q6: Win percentage seems to be on a different scale because its standard deviation is about 14, but all of the others are less than one. 

>Q7: The "0" in the chocolate column means there is no chocolate in the candy and the "1" means it contains chocolate. 

>Q8: 

```{r}
hist(candy$winpercent)
```


>Q9: The distribution of the win percentage is not symmetrical. 

>Q10: The center of the distribution is below 50%.

```{r}
chocolate_winperc <- candy$winpercent[as.logical(candy$chocolate)]
mean(chocolate_winperc)

fruity_winperc <- candy$winpercent[as.logical(candy$fruity)]
mean(fruity_winperc)
```

>Q11: On average the chocolate win percentage is higher than the fruity. 

```{r}
t.test(chocolate_winperc, fruity_winperc)
```

>Q12: According to a basic, two sample t-test, they are statistically significant with a p-value of 2.87e-8. 

```{r}
#lets try dplyr for this one
library(dplyr)
```

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
candy |>
  arrange(desc(winpercent)) |>
  head(5)
```

>Q13: The bottom 5 are: jawbusters, super bubble, chiclets, boston baked beans, and nik l nip 

>Q14: The top 5 are: reese's peanut butter cups, reese's miniatures, twix, kit kat, and snickers


>Q15: 

```{r}
library(ggplot2)
```

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "hotpink"
my_cols[rownames(candy)=="Kit Kat"] = "lightblue"
```

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```

>Q17: The worst ranked chocolate candy is Sixlets.

>Q18: The best ranked fruity candy is Starbursts. 


```{r}
library(ggrepel)
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=2.5, max.overlaps = 16)

#changed the size and overlaps so that all could be included with labels
#geom_text_repel is better than geom_text for lots of points because it reduces overlaps of words
```

>Q19: The highest win percentage for the lowest price is probably reese's minis because it has a win percent over 80%, but is close to the 25% price percentage, which is fairly low.

>Q20: The top 5 most expensive are: Nik l nip, ring pops, smarties, hershey krackel and hershey milk chocolate. The nik l nip is the least popular. 

```{r}
candy |>
  arrange(desc(pricepercent)) |>
  head(5)
```

>Q21: 

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend=reorder(rownames(candy), pricepercent), 
                   xend=0)) +
  geom_point(col=my_cols)
```


```{r}
library(corrplot) 
```

```{r}
cij <- cor(candy) 
corrplot(cij)
```

>Q22: The most anti-correlated would be chocolate and fruity (not a lot of candy is both). The pluribus and bar are also anti-correlated (which makes sense because most of the time you don't get more than one bar of candy per serving). I can tell by the size and redness of the dots. 

>Q23: Chocolate and winpercent are very correlated. Chocolate and bar are also very correlated. Same with chocolate and pricepercent, as well as bar and pricepercent. 


```{r}
noscale_pca <- prcomp(candy, scale=FALSE)
pca <- prcomp(candy, scale=TRUE)

summary(noscale_pca)
summary(pca)
```


```{r}
#change shape and colors
plot(pca$x[,1:2], col=my_cols, pch=17)
```


```{r}
candy_data <- cbind(candy, pca$x[,1:3]) 
head(candy_data)
```

```{r}
ggplot(candy_data) +
  aes(PC1, PC2, size=winpercent/100, text=rownames(candy_data), 
      label=rownames(candy_data)) +
  geom_point(col=my_cols)
```


```{r}
candy_plot <- ggplot(candy_data) +
  aes(PC1, PC2, 
      size=winpercent/100, 
      text=rownames(candy_data), 
      label=rownames(candy_data)) +
  geom_point(col=my_cols) +
  geom_text_repel(size=2.5, col=my_cols, max.overlaps = 20) +
  labs(title="Halloween Candy PCA",
       subtitle = "Colored by type: chocolate (light brown), bar (dark brown), fruity (pink), other (black)",
       caption= "data from 538")

candy_plot
```

```{r}
#library(plotly)
#ggplotly(candy_plot)
```

```{r}
barplot(pca$rotation[,1], las = 2, ylab="PC1 Contribution") 
```

>Q24: chocolate, fruity, and bar are the most "picked up." This makes sense to me because those were the categories that showed the largest pos/neg correlations on the correlation graph. So, they it makes sense that they are strongly variable. 

